\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-04-25 forwards-and-backwards.tex
theorem rewrite

\begin{definition}[Dispersal generator]
    \label{def:dispersal_generator}
    Recall that we have defined the dispersal kernel,
    $q_\theta(x, dy)$,
    to be the density of a multivariate Normal
    with mean $\meanq/\theta$ and covariance matrix $\covq/\theta$.
    This implies that if we define, for $f : \IR^d \to \IR$,
    \begin{align}
    \DG f(x)
        =
        \sum_{ij} \covq_{ij} \partial_{x_i} \partial_{x_j} f(x)
        + \sum_i \meanq_i \partial_{x_i} f(x)
    \end{align}
    then
    \begin{align}
        \theta \int \left(
            f(y) - f(x)
        \right) q_\theta(x, dy)
    \to \DG f(x) 
        \qquad \text{as } \theta \to \infty .
    \end{align}
    We will also denote the adjoint of $\DG$ by
    $$
    \DG^* f(x)
        =
        \sum_{ij} \covq_{ij} \partial_{x_i} \partial_{x_j} f(x)
        + \sum_i \partial_{x_i} (f(x) \meanq_i) .
    $$
\end{definition}

@@

\comment{
    \textbf{Note:} the previous write-up motivated things only in the time-stationary case
    (so, a stationary traveling wave, possibly with wave speed zero;
    see the corollary below).
    It'd be very nice to give this more general result,
    but the proof has to be checked to make sure it still applies,
    and the notation is time-inhomogeneous and so more annoying
}

@@

\comment{
    Allowing a more general dispersal mechanism won't just entail replacing $\Delta$ with $\DG$,
    it will also change at least one of the $\grad$s to something else (I think).
}

\begin{theorem} \label{thm:lineages}
    In the context of Definition~\ref{def:lineage_generator},
    if the dispersal process is isotropic (i.e., $\covq = \sigma^2 I$),
    then
    \begin{equation}
        \label{eqn:lineage_generator}
        \Lgen_s f
        =
        \sigma^2 r \gamma
        \left(
            \Delta f
            +
            \left(
                2 \grad \log(\gamma \varphi)
                - \meanq
            \right)
            \cdot \grad f
        \right) .
    \end{equation}
    In other words, 
    a lineage behaves as a diffusion
    that is driven by a Brownian motion 
    run at speed $\sigma^2$ multiplied by the local per-capita production of offspring ($r \gamma$)
    in a potential equal to the local birth output tilted by migraiton bias
    ($\varphi \gamma \exp(-\meanq \cdot x / 2 \sigma^2)$).
\end{theorem}

@@

@@

Writing $\hat{T}_{t,s}$ for the time-inhomogeneous semigroup
corresponding to the motion of ancestral lineages backwards in time
\comment{(Use notation for the motion of a lineage introduced above.)}
(that is, $\hat{T}_{t,s} f(y) = \int p(t,s,x,y) f(x) dy$),
we can write this as 
$$
    \int f(y)\varphi_t(y)\hat{T}_{t,s}g(y)dy
    =
    \int g(x) \varphi_s(x)
        \IE_{s,x} \left[
            \exp\left(
                \int_s^t(F+\gamma\DG r)(Y_u)du
            \right)f(Y_t)
        \right]dx.
$$
Differentiating with respect to $t$ at $t=s$, and writing 
$\hat{\mathcal L}_s$ for the generator of $\hat{T}_{t,s}$ at time $s$, we find
\begin{align*}
&
    \int f(y) \left\{
        \varphi_s(y) \widehat{\mathcal L}_s g(y)
        + \left(
            r(y) \DG^*(\gamma \varphi_s)(y) + \varphi_s(y) F(y)
        \right) g(y)
    \right\} dy \\
&\qquad =
    \int g(x) \varphi_s(x) \left(
        {\mathcal L}^Y_s f(x) + (F(x)+\gamma(x) \DG r(x)) f(x)
    \right) dx \\
&\qquad =
    \int f(x) \left(
        ({\mathcal L}^Y_s)^* (\varphi_s g)(x) + (F(x) + \gamma(x) \DG r(x)) \varphi_s(x) g(x)
    \right) dx ,
\end{align*}
where $({\mathcal L}^Y_s)^*$ is the adjoint of ${\mathcal L}^Y_s$
and we have used that $\varphi_s$ solves $\dot \varphi = r \DG^*(\gamma \varphi) + F \varphi$.
Since $f$ was arbitrary,
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{1}{\varphi_s} \left[
        ({\mathcal L}^Y_s)^* (\varphi_s g)
        + \gamma \varphi_s g \DG(r)
        - r g \DG^*(\gamma \varphi_s)
    \right] .
\end{eqnarray*}
(Note that the $\phi_s F g$ terms have cancelled.)
After liberal use of the product rule, we'll be able to simplify this last expression.
To prepare for this, first define
$$
    \DD f(x) = \sum_{ij} \covq_{ij} \partial_{ij} f(x),
$$
and note that $\DD$ satisfies the following identity:
\begin{align}
    \DD(fg)
    &=
    f \DD g
    + 2 (\covq \grad f) \cdot \grad g
    + g \DD f .
\end{align}

\comment{Note here we are allowing $\meanq$ to vary with space!}

Now consider the adjoint of $\mathcal{L}^Y_s$.
Recall that $\DG g = \DD g + \meanq \cdot \grad g$
and $\DG^* f = \DD f - \grad \cdot (f \meanq)$,
so
\begin{align*}
    ({\mathcal L}^Y_s)^* f
    &=
    r \DG^* (\gamma f) - \gamma f \DG r \\
    &=
    r \DD(\gamma f) - r \grad \cdot (\gamma f \meanq)
    - \gamma f \DG r .
\end{align*}
We can now expand the expression for $\hat{\mathcal L}_s$ to get
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{1}{\varphi_s} \left[
        r \DD(\gamma \varphi_s g) - r \grad \cdot (\gamma \varphi_s g \meanq)
        - \gamma \varphi_s g \DG(r)
        + \gamma \varphi_s g \DG(r)
        - r g \DG^*(\gamma \varphi_s)
    \right] \\
    &=&
    \frac{r}{\varphi_s} \left[
        \DD(\gamma \varphi_s g) - \grad \cdot (\gamma \varphi_s g \meanq)
        - g \DD(\gamma \varphi_s) + g \grad \cdot (\gamma \varphi_s \meanq)
    \right] \\
    &=&
    \frac{r}{\varphi_s} \left[
        \gamma \varphi_s \DD g
        + 2 (\covq \grad (\gamma \varphi_s)) \cdot \grad g
        - \gamma \varphi_s \meanq \cdot \grad g
    \right] \\
    &=&
    r \gamma \left[
        \DD g
        + 2 (\covq \grad \log(\gamma \varphi_s)) \cdot \grad g
        - \meanq \cdot \grad g
    \right] ,
\end{eqnarray*}
in agreement with \eqref{eqn:heuristic_lineage_generator}.

>>>>

\begin{definition}[Dispersal generator]
    \label{def:dispersal_generator}
    Recall that we have defined the dispersal kernel,
    $q_\theta(x, dy)$,
    to be the density of a multivariate Normal
    with mean $\meanq/\theta$ and covariance matrix $\covq/\theta$.
    This implies that if we define, for $f : \IR^d \to \IR$,
    \begin{align}
    \DG f(x)
        =
        \sum_{ij} \covq_{ij} \partial_{x_i} \partial_{x_j} f(x)
        + \sum_i \meanq_i \partial_{x_i} f(x)
    \end{align}
    then
    \begin{align}
        \theta \int \left(
            f(y) - f(x)
        \right) q_\theta(x, dy)
    \to \DG f(x) 
        \qquad \text{as } \theta \to \infty .
    \end{align}
    We will also denote the adjoint of $\DG$ by
    $$
    \DG^* f(x)
        =
        \sum_{ij} \partial_{x_i} \partial_{x_j} (\covq_{ij} f(x))
        - \sum_i \partial_{x_i} (f(x) \meanq_i) .
    $$
\end{definition}

@@

\comment{
    TODO: make notation agree with that in the proof.
}

@@

\begin{theorem} \label{thm:lineages}
    In the context of Definition~\ref{def:lineage_generator},
    \begin{align}
        \label{eqn:lineage_generator}
        \Lgen_s f
        &=
        \frac{r}{\varphi_s}
        \left[
            \DG^*(\gamma \varphi_s f) 
            - f \DG^* (\gamma \varphi_s)
        \right] \\
        &= \label{eqn:lineage_generator2}
        r\gamma
        \left[
            \sum_{ij} \covq_{ij} \partial_{ij} f
            + 2 \sum_j \left(
               \sum_i \partial_i (C_{ij} \log(\gamma \varphi_s))
            \right) \partial_j f
            - \sum_j \meanq_j \partial_j f
        \right] .
    \end{align}
\end{theorem}

@@

\begin{corollary} \label{cor:lineages_simple}
    In the context of Definition~\ref{def:lineage_generator},
    if the dispersal process is isotropic (i.e., $\covq = \sigma^2 I$),
    then
    \begin{equation}
        \Lgen_s f
        =
        \sigma^2 r \gamma
        \left(
            \Delta f
            +
            \left(
                2 \grad \log(\gamma \varphi_s)
                - \meanq
            \right)
            \cdot \grad f
        \right) .
    \end{equation}
    In other words, 
    a lineage behaves as a diffusion
    that is driven by a Brownian motion 
    run at speed $\sigma^2$ multiplied by the local per-capita production of offspring ($r \gamma$)
    in a potential equal to the local birth output tilted by migraiton bias
    ($\varphi \gamma \exp(-\meanq \cdot x / 2 \sigma^2)$).
\end{corollary}

@@

Writing $\hat{T}_{t,s}$ for the time-inhomogeneous semigroup
corresponding to the motion of ancestral lineages backwards in time
\comment{(Use notation for the motion of a lineage introduced above.)}
(that is, $\hat{T}_{t,s} f(y) = \int p(t,s,x,y) f(x) dy$),
we can write this as 
$$
    \int f(y)\varphi_t(y)\hat{T}_{t,s}g(y)dy
    =
    \int g(x) \varphi_s(x)
        \IE_{s,x} \left[
            \exp\left(
                \int_s^t(F+\gamma\DG r)(Y_u)du
            \right)f(Y_t)
        \right]dx.
$$
Differentiating with respect to $t$ at $t=s$, and writing 
$\hat{\mathcal L}_s$ for the generator of $\hat{T}_{t,s}$ at time $s$, we find
\begin{align*}
&
    \int f(y) \left\{
        \varphi_s(y) \widehat{\mathcal L}_s g(y)
        + \left(
            r(y) \DG^*(\gamma \varphi_s)(y) + \varphi_s(y) F(y)
        \right) g(y)
    \right\} dy \\
&\qquad =
    \int g(x) \varphi_s(x) \left(
        {\mathcal L}^Y_s f(x) + (F(x)+\gamma(x) \DG r(x)) f(x)
    \right) dx \\
&\qquad =
    \int f(x) \left(
        ({\mathcal L}^Y_s)^* (\varphi_s g)(x) + (F(x) + \gamma(x) \DG r(x)) \varphi_s(x) g(x)
    \right) dx ,
\end{align*}
where $({\mathcal L}^Y_s)^*$ is the adjoint of ${\mathcal L}^Y_s$
and we have used that $\varphi_s$ solves $\dot \varphi = r \DG^*(\gamma \varphi) + F \varphi$.
Since $f$ was arbitrary,
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{1}{\varphi_s} \left[
        ({\mathcal L}^Y_s)^* (\varphi_s g)
        + \gamma \varphi_s g \DG(r)
        - r g \DG^*(\gamma \varphi_s)
    \right] .
\end{eqnarray*}
(Note that the $\phi_s F g$ terms have cancelled.)
Since the adjoint of ${\mathcal L}^Y_s$ is
\begin{align*}
    ({\mathcal L}^Y_s)^* f
    &=
    r \DG^* (\gamma f) - \gamma f \DG r ,
\end{align*}
we can rewrite the generator of a lineage as 
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DG^* (\gamma \varphi_s g)
        - g \DG^*(\gamma \varphi_s)
    \right] .
\end{eqnarray*}
This is equation \eqref{eqn:lineage_generator}.

To simplify to equation \eqref{eqn:lineage_generator2},
this, first define
$
    \DD f(x) = \sum_{ij} \covq_{ij} \partial_{ij} f(x),
$
% and note that $\DD$ satisfies the following identity:
% \begin{align}
%     \DD(fg)
%     &=
%     f \DD g
%     + 2 (\covq \grad f) \cdot \grad g
%     + g \DD f .
% \end{align}
and so the adjoint of $\DD$ is
$$
    \DD^* f(x)
    =
    \sum_{ij} \partial_{ij} (\covq_{ij} f(x)) .
$$
Note that $\DD^*$ satisfies the following identity:
\begin{align}
    \DD^*(fg)
    &=
    \sum_{ij} \left\{
        g \partial_{ij} (\covq_{ij} f)
        + 2 \partial_{i} (\covq_{ij} f) \partial_j(g)
        + \covq_{ij} f \partial_{ij} g
    \right\} \\
    &=
    g \DD^* f
    + 2 \mathcal{H} f \cdot \grad g
    + f \DD g ,
\end{align}
where $(\mathcal{H} f)_j = \sum_i \partial_i (C_{ij} f)$.
    \comment{(Need better notation for that.)}
So, with $f = \gamma \varphi_s$,
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DD^*(\gamma \varphi_s g) - \grad \cdot (\gamma \varphi_s g \meanq)
        - g \DD^*(\gamma \varphi_s) + g \grad \cdot (\gamma \varphi_s \meanq)
    \right] \\
    &=&
    \frac{r}{\varphi_s} \left[
        \gamma \varphi_s \DD g
        + 2 \mathcal{H} (\gamma \varphi_s) \cdot \grad g
        - \gamma \varphi_s \meanq \cdot \grad g
    \right] \\
    &=&
    r \gamma \left[
        \DD g
        + 2 (\mathcal{H} \log(\gamma \varphi_s)) \cdot \grad g
        - \meanq \cdot \grad g
    \right] ,
\end{eqnarray*}
in agreement with \eqref{eqn:heuristic_lineage_generator}.

\begin{proof}[of Corollary \ref{cor:lineages_simple}]

In this section, we need only make
simplify equation \eqref{eqn:lineage_generator}
in the case where $\covq = \sigma^2 I$.

    \comment{TODO}
